---
title: "What makes a board game highly rated?"
output:
  github_document:
    toc: false
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.path = "figures/",
                      dpi = 150, fig.width = 8, fig.height = 4.5, fig.align = "center")
knitr::opts_knit$set(root.dir = "..")
```

**Question.** BoardGameGeek users have rated tens of thousands of games. Which features of a game, such as its length, player count, mechanics and themes, go with higher ratings? And how well can we predict a game's rating before anyone has played it?

This analysis uses the [TidyTuesday BoardGameGeek data](https://github.com/rfordatascience/tidytuesday/tree/master/data/2022/2022-01-25), cleaned by `R/prepare_data.R`, and the model functions in `R/model.R`.

```{r packages}
library(tidyverse)
library(glmnet)
library(ranger)
library(patchwork)
source("R/model.R")

ink <- "#1D1B4C"; gold <- "#F2A900"; blue <- "#2F45C9"; grey <- "#9A99B8"
theme_set(
  theme_minimal(base_size = 12) +
    theme(plot.title = element_text(face = "bold", colour = ink),
          plot.title.position = "plot",
          axis.title = element_text(colour = ink), panel.grid.minor = element_blank())
)

games  <- read_csv("data/games.csv", show_col_types = FALSE)
labels <- read_csv("data/feature_labels.csv", show_col_types = FALSE)
```

## 1. The data

After cleaning, there are **`r scales::comma(nrow(games))` games** published between 1970 and 2021, each with at least 50 user ratings. The outcome is the game's average user rating on BoardGameGeek's 1 to 10 scale.

```{r distribution, fig.height = 3.5}
ggplot(games, aes(average)) +
  geom_histogram(binwidth = 0.1, fill = ink) +
  geom_vline(xintercept = median(games$average), colour = gold, linewidth = 1.2) +
  labs(title = "Most games average between 6 and 7",
       subtitle = sprintf("Median rating %.2f (gold line)", median(games$average)),
       x = "Average user rating", y = "Games")
```

## 2. Newer games are rated much higher

```{r year-trend}
by_year <- games |>
  group_by(year) |>
  summarise(mean_rating = mean(average), n = n())

ggplot(by_year, aes(year, mean_rating)) +
  geom_point(aes(size = n), colour = ink, alpha = 0.7) +
  geom_smooth(method = "loess", se = FALSE, colour = gold, linewidth = 1.3, span = 0.4) +
  scale_size_area(max_size = 6, labels = scales::comma) +
  labs(title = "Average ratings have climbed steeply since the mid-2000s",
       x = "Year published", y = "Mean average rating", size = "Games")
```

Games from the 1990s average about 6.1, while games from 2020 average about 7.3. That is unlikely to mean games simply got that much better. Newer games are mostly rated by the enthusiasts who sought them out, while older games have been rated by a broader, more critical audience over many years. **Any model of ratings has to account for release year**, or it will mistake "recent" for "good".

## 3. Longer, more demanding games rate higher

```{r time-age, fig.height = 4}
p1 <- games |>
  mutate(time_bin = cut(playing_time, c(0, 15, 30, 60, 90, 120, 180, 240, 600),
                        labels = c("≤15", "30", "60", "90", "120", "180", "240", "240+"))) |>
  group_by(time_bin) |>
  summarise(m = mean(average), se = sd(average) / sqrt(n())) |>
  ggplot(aes(time_bin, m)) +
  geom_pointrange(aes(ymin = m - 1.96 * se, ymax = m + 1.96 * se), colour = ink) +
  labs(title = "Playing time", x = "Minutes (upper bound)", y = "Mean rating")

p2 <- games |>
  mutate(age_bin = cut(min_age, c(0, 6, 8, 10, 12, 14, 18),
                       labels = c("≤6", "8", "10", "12", "14", "18"))) |>
  group_by(age_bin) |>
  summarise(m = mean(average), se = sd(average) / sqrt(n())) |>
  ggplot(aes(age_bin, m)) +
  geom_pointrange(aes(ymin = m - 1.96 * se, ymax = m + 1.96 * se), colour = ink) +
  labs(title = "Minimum age", x = "Minimum age (upper bound)", y = NULL)

p1 + p2 + plot_annotation(title = "Games that take longer or target older players are rated higher",
                          theme = theme(plot.title = element_text(face = "bold", colour = ink)))
```

## 4. Which mechanics and themes matter, once everything else is accounted for?

Raw comparisons can mislead. Wargames, for example, tend to be long, which is itself linked to higher ratings. So I compare each feature's **raw** difference (games with it versus without it) with its **adjusted** effect from a lasso regression that holds year, length, age, player count and all other features constant.

```{r fit-models}
split  <- split_data(games)
lasso  <- fit_lasso(split$train)
forest <- fit_forest(split$train)
```

```{r mechanics, fig.height = 7.5}
raw <- labels |>
  mutate(raw_diff = map_dbl(column, \(c) mean(games$average[games[[c]] == 1]) -
                                        mean(games$average[games[[c]] == 0])),
         share = map_dbl(column, \(c) mean(games[[c]])))

co <- coef(lasso, s = "lambda.min")
adjusted <- tibble(column = rownames(co), adjusted = as.numeric(co[, 1]))

effects <- raw |>
  left_join(adjusted, by = "column") |>
  mutate(term = fct_reorder(term, adjusted))

ggplot(effects, aes(y = term)) +
  geom_vline(xintercept = 0, colour = grey) +
  geom_segment(aes(x = raw_diff, xend = adjusted, yend = term), colour = grey) +
  geom_point(aes(x = raw_diff, colour = "Raw difference"), size = 2.2) +
  geom_point(aes(x = adjusted, colour = "Adjusted (lasso)"), size = 2.8) +
  scale_colour_manual(values = c("Raw difference" = grey, "Adjusted (lasso)" = ink)) +
  facet_grid(type ~ ., scales = "free_y", space = "free_y") +
  labs(title = "Raw differences often shrink or flip once other features are held constant",
       x = "Change in average rating (points)", y = NULL, colour = NULL) +
  theme(legend.position = "top")
```

```{r top-effects}
effects |>
  arrange(desc(adjusted)) |>
  slice(c(1:5, (n() - 4):n())) |>
  transmute(Feature = as.character(term), Type = type,
            `Share of games` = scales::percent(share, 1),
            `Raw difference` = round(raw_diff, 2),
            `Adjusted effect` = round(adjusted, 2)) |>
  knitr::kable()
```

- **Roll / Spin and Move** is the clearest negative: about 0.9 points lower raw, and still 0.5 lower after adjustment, matching its reputation as a luck-driven mechanic.
- **Miniatures (+0.31) and wargames (+0.19)** keep positive effects, reflecting dedicated hobbyist audiences. **Licensed film and TV themes (−0.27)** and **humour (−0.19)** stay negative.
- **Many "modern" mechanics look great raw but not once adjusted.** Worker placement, deck building and variable player powers are each rated about half a point higher raw, but almost all of that disappears after adjustment: these games are simply newer and longer. The same is true in reverse for children's games.

## 5. Predicting a game's rating

I compare a baseline that always predicts the average rating, a lasso regression and a random forest, on a 20% held-out test set of `r scales::comma(nrow(split$test))` games.

```{r evaluation}
results <- evaluate(split$test, split$train, lasso, forest)
results |>
  mutate(rmse = round(rmse, 3), r_squared = round(r_squared, 3)) |>
  rename(Model = model, RMSE = rmse, `R²` = r_squared) |>
  knitr::kable()
```

The random forest explains about half the variation in ratings using only information available before release, and cuts the typical error from 0.84 to 0.57 rating points. The lasso is a little less accurate but fully interpretable, so it powers the interactive app.

```{r importance, fig.height = 4.5}
imp <- tibble(feature = names(forest$variable.importance),
              importance = forest$variable.importance) |>
  slice_max(importance, n = 12) |>
  left_join(labels, by = c("feature" = "column")) |>
  mutate(label = coalesce(term, recode(feature,
    year_c = "Year published", year_c2 = "Year published (curve)",
    log2_time = "Playing time", min_age = "Minimum age", max_players = "Maximum players",
    min_players = "Minimum players", n_mechanics = "Number of mechanics")),
    label = fct_reorder(label, importance))

ggplot(imp, aes(importance, label)) +
  geom_col(fill = ink) +
  labs(title = "What the random forest relies on most",
       subtitle = "Permutation importance: increase in error when a feature is shuffled",
       x = "Importance", y = NULL)
```

```{r partial-dependence, fig.height = 3.8}
pd_year <- partial_dependence(forest, split$train, "year_c", seq(-4, 1.1, by = 0.1)) |>
  mutate(value = 2010 + 10 * value)
pd_time <- partial_dependence(forest, split$train, "log2_time", log2(c(10, 15, 20, 30, 45, 60, 90, 120, 180, 240, 360))) |>
  mutate(value = 2^value)

q1 <- ggplot(pd_year, aes(value, prediction)) + geom_line(colour = ink, linewidth = 1.2) +
  labs(title = "Year published", x = NULL, y = "Predicted rating")
q2 <- ggplot(pd_time, aes(value, prediction)) + geom_line(colour = ink, linewidth = 1.2) +
  scale_x_log10(breaks = c(10, 30, 60, 120, 360)) +
  labs(title = "Playing time (minutes, log scale)", x = NULL, y = NULL)
q1 + q2 + plot_annotation(title = "Holding everything else fixed, recency and length both raise predicted ratings",
                          theme = theme(plot.title = element_text(face = "bold", colour = ink)))
```

## 6. Conclusions

- **Release year is the single strongest predictor**, most likely because recent games are rated by self-selected enthusiasts. Comparing games from different eras on raw ratings is unfair to older titles.
- **Longer, more demanding games rate higher**, with ratings rising steadily with playing time and recommended age.
- **Mechanics matter, but less than raw averages suggest.** After adjustment, roll-and-move is the clearest negative and miniatures, wargames and solo play the clearest positives.
- **Ratings are partly predictable before release**: a random forest explains about half the variation.

**Limitations.** These are associations, not causal effects: adding miniatures to a game will not by itself raise its rating. BoardGameGeek raters are a self-selected hobbyist community, so the results describe what *that* audience likes. Only the 25 most common mechanics and 20 most common categories are modelled, and game complexity (BGG's "weight") is not in this dataset.

```{r export, include = FALSE}
export_app_files(lasso, games, labels, dir = "app")
write_csv(results, "analysis/model_results.csv")
```
